---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

Load the normalized counts, the targets and the GFP counts.

```{r}
library(tidyverse, quietly = TRUE, verbose = FALSE)
setwd("E:/Xuexi/miRNA_footprint/HEK_Rzepiela")
```

```{r}

counts = readRDS(file = "./Data/counts.rds")
targets = readRDS(file = "./Data/Target_files/Targets__combined_199-5p_199-3p.rds")
GFP_counts = readRDS(file = "./Data/GFP_counts.rds")

```

Calculate the activity. The activity is calculated as the log2 of the normalized counts divided by the mean of the normalized counts. 

Strong miRNA activity → Lower target gene expression → Higher activity score
Weak miRNA activity → Higher target gene expression → Lower activity score

```{r}
calculate_activity = function(counts, targets) {
  cell_sums = counts[targets$ensembl_gene_id,] %>% na.omit() %>% colSums() %>% as.data.frame()
  cell_sums_norm = cell_sums/mean(cell_sums$.)
  activity = -log2(cell_sums_norm) + 1
  return(activity)
}

activity = calculate_activity(counts = counts, targets = targets)
```

Find the correlation between the miRNA activity and the GFP counts.


```{r}
cor_df = merge(activity, GFP_counts, by.x = 0, by.y = 0); rownames(cor_df) = cor_df$Row.names; cor_df = cor_df[,2:3];colnames(cor_df) = c("activity", "GFP")
cor_df = cor_df[cor_df$GFP < 20000,]
cor = cor.test(cor_df$activity, cor_df$GFP, method = "spearman", exact = FALSE)
```
```{r}
cor2 = cor.test(cor_df$activity, cor_df$GFP, method = "pearson", exact = FALSE)
```

The dot density for plotting the correlation. The density is calculated using the kde2d function from the MASS package.

```{r}
get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}
# n controls the grid size or smoothing. Higher n = finer grain density estimation
cor_df$density = get_density(cor_df$activity, cor_df$GFP, n = 100)
```

Plot the correlation

```{r}
library(viridis)
ggplot() + geom_point(data = cor_df,aes(activity, GFP, color = density))  + scale_color_viridis() +
  geom_smooth(data = cor_df,aes(activity, GFP),method = "lm", se = TRUE, colour = "black") +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = paste0("Correlation   $\\rho$ = ", round(cor[4]$estimate, digits = 3),"  p-value = ", formatC(cor[3]$p.value,format = "e", digits = 3  )  )) +
  theme_bw(base_size = 16)
```

Show the top 5% vs bottom 5%

```{r}
top = arrange(cor_df, desc(activity))[1:164, ]
top$Type = "top"
mean_GFP_top = mean(top$GFP)

bottom = arrange(cor_df, activity)[1:164, ]
bottom$Type = "bottom"
mean_eGFP_bottom = mean(bottom$GFP)

plot_5percent = rbind(top,bottom)
```


```{r}
ggplot(plot_5percent) + geom_violin(aes(x = Type, y = GFP, fill = Type)) +
  ggtitle("Top 5% vs bottom 5% comparison") +
  labs(x = "") + 
  scale_fill_manual(values=c("palevioletred2", "palegreen3")) +
  theme(legend.position = "None", 
        plot.title = element_text(hjust = 0.5), 
        panel.background = element_blank(),
        axis.line = element_line()) +
    stat_summary(aes(x = Type, y = GFP, fill = Type),fun = "mean",
               geom = "point",
               color = "black")
```


```{r}
ggplot(plot_5percent) + 
  # Add violin plot with transparency
  geom_violin(aes(x = Type, y = GFP, fill = Type), 
             alpha = 0.7,
             trim = FALSE) +
  
  # Add boxplot inside violin for quartile information
  geom_boxplot(aes(x = Type, y = GFP), 
               width = 0.15,
               fill = "white",
               alpha = 0.5) +
  
  # Add individual points with jitter
  #geom_jitter(aes(x = Type, y = GFP),
  #            width = 0.1,
  #            alpha = 0.4,
  #            size = 1) +
  
  # Add mean point with custom shape and size
  stat_summary(aes(x = Type, y = GFP),
              fun = mean,
              geom = "point",
              size = 3,
              color = "black",
              shape = 23,
              fill = "white") +
  
  # Add error bars for standard error
  stat_summary(aes(x = Type, y = GFP),
              fun.data = mean_se,
              geom = "errorbar",
              width = 0.1) +
  
  # Enhanced title and labels
  ggtitle("Top 5% vs Bottom 5% GFP Comparison") +
  labs(x = "",
       y = "GFP Expression") +
  
  # Custom colors with slightly muted tones
  scale_fill_manual(values = c("#FF9DB5", "#90EE90")) +
  
  # Enhanced theme
  theme_minimal() +
  theme(
    # Text elements
    plot.title = element_text(hjust = 0.5,
                            face = "bold",
                            size = 14,
                            margin = margin(b = 20)),
    axis.title.y = element_text(size = 12,
                               margin = margin(r = 10)),
    axis.text = element_text(size = 10),
    
    # Plot elements
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(fill = NA, color = "gray80"),
    axis.line = element_line(color = "black"),
    
    # Legend
    legend.position = "none"
  )
```

Here we exclude cells based on their read depth. Check how the correlation changes depending on the cutoff.
```{r}
raw_counts = readRDS("./Data/raw_counts.rds")
```

```{r}


raw_sums = colSums(raw_counts) %>% as.data.frame()

x = 2000
df = lapply(seq(0,60000,10), function(x){
  top_cells = raw_counts[,raw_sums > x] %>% colnames()
  activity = calculate_activity(counts = counts[,top_cells], targets = targets)
  colnames(activity) <- "activity"
  a <- merge(activity,GFP_counts, by.x = 0, by.y = 0)
  cor <- cor.test(a$activity, a$eGFP, method = "spearman", exact = FALSE)
  return(c(cor$estimate, x))
}) %>% as.data.frame()

d = df %>% t() %>% as.data.frame(); colnames(d) = c("rho", "cutoff"); rownames(d) = seq(1,dim(d)[1])

p1 = ggplot() + geom_histogram(data = raw_sums, aes(x = .), alpha = 0.5, color = "black", fill = "grey90") + 
  geom_point(data = d, aes(x = cutoff, y = rho*1000), color = "cornflowerblue", size = 1.5) +
   theme_classic() + xlab(label = "Raw count sum") + scale_y_continuous(
    name = "Freq",
    sec.axis = sec_axis( trans=~./1000, name="spearman rho"))
  

p1
```

```{r}
best_cor = slice_max(.data = d, order_by = rho)$rho
best_cor_cutoff = slice_max(.data = d, order_by = rho)$cutoff # 48630
best_cells = raw_sums %>% filter(. > best_cor_cutoff) %>% rownames()
```

Find the best correlation and plot at that point.

```{r}
sub_cor_df = cor_df[best_cells,]
cor_sub =  cor.test(sub_cor_df$activity, sub_cor_df$GFP, method = "spearman", exact = FALSE)

ggplot() + geom_point(data = sub_cor_df,aes(activity, GFP),color = "cornflowerblue") +
  geom_smooth(data = sub_cor_df,aes(activity, GFP),method = "lm", se = TRUE, colour = "black") +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = paste0("hsa-mir-199-3p   $\\rho$ = ", round(cor_sub[4]$estimate, digits = 3),"  p-value = ", formatC(cor_sub[3]$p.value,format = "e", digits = 3  )  )) +
  theme_bw(base_size = 16)

```
